---
title: "Cyclistic data analysis - analyzing"
author: "Oscar Fernandez Solano"
date: "2024-02-12"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

After concluding the cleaning phase ([part 1](https://github.com/OscarFernandezMX) and [part 2](https://github.com/OscarFernandezMX)), we can start analyzing the data.

We use the following libraries.
```{r libraries}
library(tidyverse)
library(skimr)
library(hms)
library(scales)
```

Next we set the working directory, the data path, and read the proper csv file.
```{r setting working directory and reading data}
setwd("~/GitHub/Data_Science_Projects")
data_path = "Data/cyclistic/"
file_name = paste0(data_path, "2023_cyclistic_tripdata_downsampled.csv")
# There is a problem when exporting the data to csv where some values for ride length are missing. Therefore, we create this row again after loading the data.
# When reading the days, the format has changed, therefore we create the day_of_week column again.
df <- read_csv(file_name) %>%
  mutate(
    ride_length = as_hms(ended_at - started_at),
    day_of_week = format(as.Date(started_at), "%A")
    )
  
skim_without_charts(df)
```


### Separating date in day, month, and year

For our analysis, we might want to check for trends based on the month, as well as the season. Therefore, we need to extract the month from the starting dates. We'll also add the year and day columns.
```{r day, month, and year}
df <- df %>%
  mutate(
    day = format(as.Date(started_at), "%d"),
    month = format(as.Date(started_at), "%m"),
    year = format(as.Date(started_at), "%Y")
  )

head(df)
```


### Selecting attributes of interest

For this part of the analysis, we focus more on the rides, rather than their location. Therefore, we select these attributes for this part and create a new dataframe ignoring the other attributes related to location (latitude and logitude).

First we see the 14 attributes from our data plus the 4 added from the above steps.
```{r all attributes}
colnames(df)
```

Now we select the attributes of interest.
```{r selecting attributes}
rides <- df %>%
  select(ride_id, rideable_type, started_at, ended_at, day, month, day_of_week, ride_length, start_station_name, member_casual) %>%
  arrange(started_at)

head(rides)
```
With this data we can start our analysis now.


## Descriptive analysis

The first part is to summarize the data. We'll do it by day of the week and by month. But first let's take a look at the whole data.

We can see the following statistics for the `ride_length` attribute. First we can see the minimum, 1st quartile (25th percentile), median (also the 2nd quartile), 3rd quartile (75th percentile), mean, and maximum time.
```{r summary ride length}
rides %>%
  summarize(t_min = min(ride_length),
            t_q1 = quantile(ride_length, probs = 0.25),
            t_median = median(ride_length),
            t_q3 = quantile(ride_length, probs = 0.75),
            t_mean = mean(ride_length),
            t_max = max(ride_length))
```

Next we show the same statistics for each member type.
```{r stats for ride_length per member_casual}
rides %>%
  group_by(member_casual) %>%
  summarize(
    t_min = min(ride_length),
    t_q1 = quantile(ride_length, probs = 0.25),
    t_median = median(ride_length),
    t_q3 = quantile(ride_length, probs = 0.75),
    t_mean = as_hms(round(mean(ride_length), 2)),
    t_max = as_hms(round(max(ride_length), 2))
  )
```


### By day of the week

We can now summarize the data by day of the week. This code chunck shows the same statistics from the beginning, but now it is grouped by day of the week and member type.
```{r stats by day_of_week}
summary_by_day_of_week <- rides %>%
  mutate(day_of_week = ordered(day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>%
  group_by(day_of_week, member_casual) %>%
  summarize(
    number_of_rides = n(),
    t_min = min(ride_length),
    t_q1 = quantile(ride_length, probs = 0.25),
    t_median = median(ride_length),
    t_q3 = quantile(ride_length, probs = 0.75),
    t_mean = as_hms(round(mean(ride_length), 2)),
    t_max = as_hms(max(ride_length))
  )

summary_by_day_of_week
```

With this data we can make some plots using only the number of rides and the mean ride duration.
```{r plot number of rides per day}
ggplot(data = summary_by_day_of_week,
       (aes(x = day_of_week,y = number_of_rides, fill = member_casual))) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides", subtitle = "per day and member type") +
  xlab("Day of the week") +
  ylab("Number of rides") +
  scale_y_continuous(labels = unit_format(unit = "K", scale = 1e-3))
```
This graph shows that casual users tend to ride more than members on weekends. Member users ride more during the weekdays, and the numbers are very similar each workday. On weekends, specially Saturday and Sunday, they ride there are less rides than usual.

```{r plot average ride duration per day}
ggplot(data = summary_by_day_of_week,
       aes(x = day_of_week, y = t_mean, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average ride duration", subtitle = "per day and member type") +
  xlab("Day of the week") +
  ylab("Average ride duration")
```
The average duration of casual users' rides is almost twice average from member rides. In both cases, rides tend to last longer during the weekend, but this trend is more visible on casual members.


### By month

Another interesting thing, is to observe this data by month. Therefore we group the data and show the same statistics based on the month (recall that this data is for the year 2023).
```{r stats by month}
summary_by_month <- rides %>%
  group_by(month, member_casual) %>%
  summarize(
    number_of_rides = n(),
    t_min = min(ride_length),
    t_q1 = quantile(ride_length, probs = 0.25),
    t_median = median(ride_length),
    t_q3 = quantile(ride_length, probs = 0.75),
    t_mean = as_hms(round(mean(ride_length), 2)),
    t_max = as_hms(max(ride_length))
  )
summary_by_month
```

With this data we can make similar plots.
```{r plot number of rides per month}
ggplot(data = summary_by_month,
       aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides", subtitle = "per month and member type") +
  xlab("Month") +
  ylab("Number of rides") +
  scale_y_continuous(labels = unit_format(unit = "K", scale = 1e-3))
```
We can see that there is a considerable increase in the use of the bicycle starting from March and peaking at August. In September, the use starts to decrease and by December it is very low, specially in the causal riders. In these months, the member riders use the bicycle more than casual members. Casual users ride more than members from May to September.

```{r plot average ride duration per month}
ggplot(data = summary_by_month,
       aes(x = month, y = t_mean, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average ride duration", subtitle = "per month and member type") +
  xlab("Month") +
  ylab("Average ride duration")
```
This trend also affects the average ride duration. People tend to ride more time from April, peaking at July, then decreasing smoothly. Until November, the time average time is still higher than 20 minutes. However, this trend is more smooth for members, where the average ride duration is almost the same, no matter the month. Also, we can notably see that casual members ride more time than members each month.


### Distribution of start time

From the data we can see that the members are more consistent with the use of the bicycle. This might be because members probably use the bike to commute to work. In this step we are going to visualize the distribution of start and end hour of the rides to investigate this trend further.
```{r distribution of hours}
hours_distribution <- rides %>%
  mutate(day_of_week = ordered(day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>%
  mutate(
    started_at = as_hms(format(started_at, "%H:%M:%S")),
    ended_at = as_hms(format(ended_at, "%H:%M:%S"))
  )
```

#### By day

```{r start rides distribution per day for members}
hours_distribution %>%
  filter(member_casual == "member") %>%
  ggplot(aes(x = started_at)) +
  geom_histogram(bins = 24) +
  facet_wrap(~day_of_week) +
  labs(title = "Distribution of start times for members", subtitle = "by day of the week") +
  xlab("The hour at which the ride started") +
  ylab("Number of rides")
```
In this plot we have some evidence that could suggest that members indeed use the bicycle to commute to work. Since we can see some peaks before 9:00 and after 15:00 h, which is within the workday. Interestingly, on weekends we can't see that trend, because most people work from Monday to Friday.

```{r start rides distribution per day for casual users}
hours_distribution %>%
  filter(member_casual == "casual") %>%
  ggplot(aes(x = started_at)) +
  geom_histogram(bins = 24) +
  facet_wrap(~day_of_week) +
  labs(title = "Distribution of start times for casual users", subtitle = "by day of the week") +
  xlab("The hour at which the ride started") +
  ylab("Number of rides")
```
In the case of casual users, we can also see two main peaks, but the highest is after 15:00 h. While this might also suggest that causal riders commute on the bike to work, the available data is not enough to support more assumptions. The only thing we might suspect is that, since casual users ride more time, they might ride after work, not precisely to commute to their homes, but as a recreation ride.

#### By month

```{r start rides distribution per month for members}
hours_distribution %>%
  filter(member_casual == "member") %>%
  ggplot(aes(x = started_at)) +
  geom_histogram(bins = 24) +
  facet_wrap(~month)
```

```{r start rides distribution per month for casual users}
hours_distribution %>%
  filter(member_casual == "casual") %>%
  ggplot(aes(x = started_at)) +
  geom_histogram(bins = 24) +
  facet_wrap(~month)
```
We can see that in both cases, the trend is the same for the two type of users for each month, where we can only see the difference in the amount of rides.


## Conclusions from the analysis

Based on the available data, we can see two important things:  
- It is probable that casual riders use the bicycle more for recreational uses. This explains why they ride more time, and the seasonal use (see graph below - it's not very enjoyable to ride in low temperatures).  
- Probably, people with membership use the bicycle to commute to work. This might explain why their rides last less and keep almost the same during the weekdays, and in every month. Though in winter they tend to ride less, those who ride, still use the bicycle to commute, therefore, they take almost the same time to arrive, which might explain why the average time remains almost the same every month.

```{r plot number of rides per month with annotation}
ggplot(data = summary_by_month,
       aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides", subtitle = "per month and member type") +
  xlab("Month") +
  ylab("Number of rides") +
  scale_y_continuous(labels = unit_format(unit = "K", scale = 1e-3)) +
  annotate("rect", xmin=c(0, 11.5), xmax=c(3.5, 13),
           ymin=c(0) , ymax=c(250000), alpha=0.1, color="blue", fill="blue") +
  annotate("text", x = 1.5, y = 200000, 
           label = "Winter", color="orange", 
           size = 8, angle = -15, fontface="bold")
```






















