---
title: "Cyclistic data analysis - cleaning"
author: "Oscar Fernandez Solano"
date: "2024-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The first part of the cleaning phase can be found [here](https://github.com/OscarFernandezMX). In the first part we used Apache Spark and SQL to clean the data. Here is a summary of the steps made:
- Inspected each row for inconsistencies.
- Removed duplicates.
- Removed rows containing NA in stations' information.

As mentioned in the pervious phase, we need to make a downsampling to have balanced classes. This is because we want to find differences between casual cyclists and members.

We use the `tidyverse`library.
```{r libraries}
library(tidyverse)
library(skimr)
library(janitor)
```

Next we set the working directory, the data path, and read the csv file created from the preparing phase.
```{r setting working directory}
setwd("/Users/oscarfs/GitHub/Data_Science_Projects")
data_path = "Data/cyclistic/"
file_name = paste0(data_path, "2023_cyclistic_tripdata_clean.csv")
df <- read_csv(file_name)
```

We get the summary of the data. From this summary, we can notice what 
```{r}
glimpse(df)
```

```{r}
skim_without_charts(df)
```















First, we save the data from the "casual" class.
```{r}
casual_df <- df %>%
  filter(member_casual == "casual")
```

Now we randomly sample cases with "member" class.
```{r}
downsampled_df <- df %>%
  filter(member_casual == "member") %>%
  sample_n(size = nrow(casual_df), replace = FALSE, seed = 123)
```

Finally, we can bind both dataframes.
```{r}
downsampled_df <- bind_rows(casual_df, downsampled_df)

file_name = paste0(data_path, "2023_cyclistic_tripdata.csv")

if (!file.exists(file_name)) {
  df_files = list()
  df <- bind_rows(lapply(files, read_csv))
  write_csv(df, paste0(data_path, "2023_cyclistic_tripdata.csv"))
  print("File created.")
```